Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-16 introduction.tex
Added sky localization uncertainty table
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1081 ccdd05d5-56c1-461a-bf98-85f6876523b4

It has been observed that by examining the signal to noise ratio (\SNR{}) stream for
threshold crossings before the \GW{} signal leaves the detection band, it is
possible to trade some \SNR{} and sky localization accuracy for pre-merger
detection. Figure~\ref{fig:earlywarning} shows projected early detection rates
for \textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming event rates
predicted in \citet{Abadie:2010p10836} and anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise}.
The most realistic estimates indicate that at an \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$, but $\sim$10~yr$^{-1}$ will be
detectable within 10~s of merger and $\sim$1~yr$^{-1}$ will be detectable
within 100~s of merger. These rates are shown with their substantial uncertainties in figure~\ref{fig:earlywarning}.
%
\begin{figure}[h]
\plotone{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that will be detectable by Advanced \LIGO\ $t$ seconds before coalescence.
The heavy solid line is the most realistic yearly rate estimate.  The shaded
region represents the 5 to 95\% confidence interval arising from uncertainty
in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}

Telescopic followup requires estimating the location of the \GW{} source. The
localization uncertainty is dominated by uncertainty in the time of arrival of
the \GW{}s, which is determined by the signal's effective bandwidth and \SNR{}
\citep{Fairhurst2009}. Unfortunately, for a \GW{} signal with an \SNR{} of
blah, the 90\% confidence area is at least blah deg$^2$ (blah\% of the sky) at
10~s before merger and blah deg$^2$ (blah\% of the sky) at 1~s before merger.
It is infeasible to search this large an area for a prompt counterpart. For
reference, \textsc{lsst} should see first light around the same time as
Advanced \LIGO{} and will have an unparalleled field of view of 9.6 deg$^2$
\citep{2008arXiv0805.2366I}.

If we can take galactic luminosities to accurately trace mass and thus proxy
for the \CBC{} event rate, it is possible to reduce the localization
uncertainty using a galaxy catalog as a prior probability distribution
\citep{galaxy-catalog}, along with the sky location and luminosity distance
estimate from the \GW{} signal, as was done in S6/VSR3. At Advanced \LIGO{}
\textsc{ns}--\textsc{ns} horizon distances of $\sim$400~Mpc, the galaxy catalog
is fairly isotropic and thus not as useful for downselecting pointings, but
sources that are detectable before merger are likely to be exceptionally
nearby. Thus there will be relatively few galaxies to consider for a given
early-warning candidate and the galaxy catalog is likely to be sufficiently
complete. This should allow us to shrink the 90\% confidence area substantially.
Even so, it will be desirable to mobilize a large number of heterogeneous
telescopes in order to guarantee imaging this exceptional event at the instant
of merger. Because the sky localization will improve rapidly down to blah
deg$^2$ at merger, by the time that the entire \CBC{} event is over it will be
possible to send a second alert to downselect the images that have already been
taken as well as repoint telescopes to more likely locations.

>>>>

It has been observed that by examining the signal to noise ratio (\SNR{}) stream for
threshold crossings before the \GW{} signal leaves the detection band, it is
possible to trade some \SNR{} and sky localization accuracy for pre-merger
detection. Figure~\ref{fig:earlywarning} shows projected early detection rates
for \textsc{ns}--\textsc{ns} binaries in Advanced \LIGO\ assuming event rates
predicted in \citet{Abadie:2010p10836} and anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise}.
The most realistic estimates indicate that at an \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$, but $\sim$10~yr$^{-1}$ will be
detectable within 10~s of merger and $\sim$5~yr$^{-1}$ will be detectable
within 25~s of merger. These rates are shown with their substantial uncertainties in figure~\ref{fig:earlywarning}.
%
\begin{table}
\caption{\label{table:sky-localization-accuracy}Horizon distance, most likely yearly detection rate, and \SNR\ at merger for a source that produces an \SNR\ of 8 in all three detectors at $t$~s before merger.  The fifth and sixth columns show the minimum area of the 90\% confidence region on the sky before merger and at merger.}
\begin{tabular}{rrrrrr}
\tableline\tableline
& horiz. & rate & final & \multicolumn{2}{c}{$A$(90\%) (deg$^2$)} \\
\cline{5-6}
$t$ (s) & (Mpc) & (yr$^{-1}$) & \SNR\ & early & final \\
\tableline
25.0 & 203 & 5 & 15.8 & 2893.2 & 1.8 \\
10.0 & 269 & 12 & 11.9 & 1150.6 & 3.6 \\
5.0 & 307 & 18 & 10.4 & 594.6 & 5.1 \\
1.0 & 360 & 29 & 8.9 & 153.9 & 7.5 \\
0.1 & 391 & 37 & 8.2 & 32.5 & 9.1 \\
0.0 & 400 & 40 & 8.0 & 9.6 & 9.6 \\
\tableline
\end{tabular}
\end{table}
%
\begin{figure}[h]
\plotone{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning}Expected number of \textsc{ns}--\textsc{ns}
sources that will be detectable by Advanced \LIGO\ $t$ seconds before coalescence.
The heavy solid line is the most realistic yearly rate estimate.  The shaded
region represents the 5 to 95\% confidence interval arising from uncertainty
in predicted event rates \citep{Abadie:2010p10836}.}
\end{figure}

Telescopic followup requires estimating the location of the \GW{} source. The
localization uncertainty is dominated by uncertainty in the time of arrival of
the \GW{}s, which is determined by the signal's effective bandwidth and \SNR{}
\citep{Fairhurst2009}.  Table~\ref{table:sky-localization-accuracy} shows the
minimum 90\% confidence area for a source that produces an \SNR\ of 8 in the
\LIGO\ detectors in Hanford, Washington, and Livingston, Louisiana, and, the
Virgo detector in Cascina, Italy, at a time $t$ before merger.  Also shown are
the final \SNR\ and 90\% confidence area at merger for the same source.
At 10~s before merger, the pre-merger 90\% confidence area is about $1150.6$~deg$^2$
(2.8\% of the sky) and at 1~s before merger is 153.9~deg$^2$ (0.37\% of the sky).
The confidence region shrinks quadratically in area as the overall amplitude of the
signal increases.

It is infeasible to search this large an area for a prompt counterpart. For
reference, \textsc{lsst} should see first light around the same time as
Advanced \LIGO{} and will have an unparalleled field of view of 9.6 deg$^2$
\citep{2008arXiv0805.2366I}.  However, if we can take galactic luminosities
to accurately trace mass and thus proxy
for the \CBC{} event rate, it is possible to reduce the localization
uncertainty using a galaxy catalog as a prior probability distribution
\citep{galaxy-catalog}, along with the sky location and luminosity distance
estimate from the \GW{} signal, as was done in S6/VSR3. At Advanced \LIGO{}
\textsc{ns}--\textsc{ns} horizon distances of $\sim$400~Mpc, the galaxy catalog
is fairly isotropic and thus not as useful for downselecting pointings, but
sources that are detectable before merger are likely to be exceptionally
nearby. Thus there will be relatively few galaxies to consider for a given
early-warning candidate and the galaxy catalog is likely to be sufficiently
complete. This should allow us to shrink the 90\% confidence area substantially.
Even so, it will be desirable to mobilize a large number of heterogeneous
telescopes in order to guarantee imaging this exceptional event at the instant
of merger. Because the sky localization will improve rapidly until merger,
by the time that the entire \CBC{} event is over it will be
possible to send a second alert to downselect the images that have already been
taken as well as repoint telescopes to more likely locations.
